---
title: "EDA"
format: html
editor: visual
---

## Introduction

The data that I will be working with is the Diabetes Health Indicators Dataset. This is a cleaned dataset of 253,680 responses to the CDC's Behavioral Risk Factor Surveillance System (BRFSS), a system of health-related telephone surveys. The data includes the response variable 'Diabetes_binary' indicating that a respondent either has diabetes or prediabetes [1], or does not have diabetes [0]. Twenty-one other variables are available that include both qualitative and quantitative variables. This data is commonly used to answer questions such as "which variables are accurate predictors of whether an individual has diabetes." In addition to the response variable 'Diabetes_binary' I will be using the following predictor variables for my analysis:

- Smoker: Have you smoked at least 100 cigarettes in your entire life [0 = no, 1 = yes]

- GenHlth: Would you say your general health is [1 = excellent, 2 = very good, 3 = good, 4 = fair, 5 = poor]

- Sex: Gender [0 = female, 1 = male]

- Age: 13 level age category [1 = 18-24, 9 = 60-64, 13 = 80 and older]

- Income: 8 level income scale [1 = <10,000, 5 = <35,000, 8 = >75,000]


The goal of exploratory data analysis is to examine the underlying data. This is done by making sure each variable is assigned their appropriate category (dlb, char, factor), examining numeric summaries of the variables, and exploring different plots to visually inspect the data. Another aspect to check is missing values, as this can impact results calculated from the data. We want to make sure we have clean data so that our modeling can be as accurate as possible. For this project, our goal with modeling is to create an accurate representation of the data contained in the given set. We want to utilize the best possible model so that we can make predictions about future outcomes and identify underlying patterns. To achieve this, we train parameters and explore different types of models to get best representation of the data. 


## Data

Reading in the diabetes dataset. 
```{r}
library(tidyverse)

diab_data <- read_csv("diabetes_binary_health_indicators_BRFSS2015.csv")
diab_data
```

Subsetting the data, converting variables where necessary, and assigning appropriate labels.
```{r}
diab_data <- diab_data |>
  mutate(diab_bin = factor(Diabetes_binary, levels = c(0, 1),
                           labels = c("no diabetes", "diabetes")),
         smoker = factor(Smoker, levels = c(0, 1), 
                         labels = c("no", "yes")),
         genHlth = factor(GenHlth, levels  = c(1, 2, 3, 4, 5), 
                          labels = c("excellent", "very good", "good", "fair", "poor")),
         sex = factor(Sex, levels = c(0, 1), 
                      labels = c("female", "male")),
         age = factor(Age, levels = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13), 
                      labels = c("18-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54", "55-59", "60-64", "65-69", "70-74", "75-79", "80+")),
         income = factor(Income, levels = c(1, 2, 3, 4, 5, 6, 7, 8),
                         labels = c("<10", "10-15", "15-20", "20-25", "25-35", "35-50", "50-75", "75+"))) |>
  select(diab_bin, smoker, genHlth, sex, age, income) 
diab_data
```

Investigating missing values, there do not appear to be any in the variables which makes sense because the owener of the dataset removed them.
```{r}
sum_na <- function(column){
 sum(is.na(column))
}
na_counts <- diab_data |>
 summarize(across(everything(), sum_na))

na_counts
```

## Summarizations

** Might want to split up
Looking into contingency tables to get counts of different groupings of variables. Looking through these values we notice some interesting trends relating the response variable to general health and age. The gap between diabetes and no diabetes numbers gets smaller as health score declines. With the age variable we notice increasing numbers of responses with diabetes as age increases. 
```{r}
diab_ct <- table(diab_data$diab_bin)
diab_ct

diab_smoker <- table(diab_data$diab_bin, diab_data$smoker)
diab_smoker

diab_hlth <- table(diab_data$diab_bin, diab_data$genHlth)
diab_hlth

diab_sex <- table(diab_data$diab_bin, diab_data$sex)
diab_sex

diab_age <- table(diab_data$diab_bin, diab_data$age)
diab_age

diab_inc <- table(diab_data$diab_bin, diab_data$income)
diab_inc
```

In this plot we see the number of responses that have diabetes or not grouped by their general health classification split between males and females. We see that the groups withing the has diabetes category appear similar between males and females
```{r}
ggplot(diab_data, aes(x = diab_bin)) +
  geom_bar(aes(color = genHlth)) +
  facet_wrap(~sex)
```

Similarly, here we see the breakdown of income between smokers and non smokers to be relatively similar between people with and without diabetes.
```{r}
ggplot(diab_data, aes(x = diab_bin)) +
  geom_bar(aes(color = income)) +
  facet_wrap(~smoker)
```


[Click here for the modeling page](Modeling.html) 